The Forecast Trap

### 2022-04-12 manuscript/round4-manuscript.Rmd

- abbreviated running title: "The Forecast Trap"
- Authorship statement: _This is a single-author paper_
- data accessibility statement: _All simulation data generated for analyses here, along with code required for the analysis, is available in the Zenodo Data archive, https://doi.org/10.5281/zenodo.4660621_
- number of words in the abstract: 148
- number of words in main text: 4900
- number of cited references: 44
- number of tables & figures: 5 Figures

@@

In this paper, I demonstrate that the model that makes the better prediction (defined as a strictly proper score, @Gneiting2007) is not necessarily the model that makes the better policy (defined in terms of utility, e.g. expected net present value, @Clark1990).
I show that our best methods for learning about model structure or parameters by repeatedly comparing forecasts to observations can be counter-productive.
Put another way, the value of information (VOI, as measured by the expected utility given that information minus the utility without it; see @Howard1966; @Katz1987), can actually be negative.
When VOI is negative, the decision-maker may become trapped into accepting mediocre outcomes derived from a model that makes accurate forecasts, even when a less accurate model that would generate better outcomes is available.
I will present two examples of this "forecast trap" and examine how it arises as a result of non-uniqueness of models [@Oreskes1994; @Schindler2015] with respect to either of these objectives.

@@

I shall focus on examples involving fisheries management to illustrate principles shared in many ecological systems.
Fisheries are a significant economic and conservation concern worldwide and their management remains an important debate [e.g. @Worm2006; @Worm2009; @Costello2016].
Moreover, their management has been a proving grounds for theoretical and practical decision-making issues [e.g. @Clark1973; @Reed1979; @Walters1981; @Ludwig1982] arising in a wide range of other contexts, including
invasive species [@Boettiger2021], infectious diseases [@Li2017; Shea2014], fire management [@Richards1999] conservation planning and priortization [@Chades2008; @Wilson2006] climate policy [@] and much else [@Ludwig1993; @Lande1994; @Polasky2011].

@@

To better understand how a model can produce a more accurate forecast and yet still lead to a worse decision, it may be helpful to start with a concrete example.
Fig 1 shows both the forecasts and realized management outcomes of two alternative three-species models, "A" and "B" (see Appendix A) in predicting the population dynamics of striped bass (an economically important fishery) and double-crested cormorants (a target species for conservation) which both feed on a population of river herring (whose abundance we assume is not measured). 
Model A accurately forecasts the abundance of both bass and cormorants well into the future, but the optimal management strategy derived from its forecasts leads to steadily declining species abundances and overall disappointing net utility.
Model B produces substantially less accurate forecasts, but nevertheless achieves better outcomes.
The net utility under model B is in fact nearly identical to the maximum utility attainable given the true model, while management under model A achieves only 38% of that utility.

@@

Faced with a collection of models, a manager can seek either to identify the best model to use, or to consider an integrative assessment which uses the whole ensemble of models to represent the manager's uncertainty about the underlying process.
I will consider both approaches in turn. 
Figure 2 compares forecasts generated by two of the candidate models to observations drawn from simulations of the underlying process.
As before, these are true forecasts: the model forecasts are generated first, they have not been fit to these observations.
In the un-fished scenario (top panels), both models try to predicting the same un-fished equilibrium dynamics.
In the second scenario (lower panel), the manager uses the optimal SDP policy derived from each forecast to determine the TAC for the following year, and compares the observed stock size to that which the model predicted given that fishing quota.
In both cases, model 2 provides far more accurate forecasts, as seen in the error bars and confirmed by the distribution of proper scores [Fig 3C-D; @Gneiting2007].
Model and simulation details are provided in Appendix B.

@@

Despite the clearly superior predictive accuracy of model 2 in both scenarios, the outcomes from management under model 2 are substantially worse.
We can assess such outcomes in less abstract terms than forecasting skill, such as economic value them manager sought to optimize (in dollars) or the ecological value (unharvested biomass) [Fig 4].

@@

A manager operating under model 2 would have little indication that the model was flawed: both future stock sizes and expected harvest yields consistently match model predictions.
This manager would be stuck in the forecast trap, incorrectly mistaking the a degraded ecological state and reduced economic outcomes as the best that can be achieved because future observations continue to validate this model.
Had we been able to include Model 3 in our forecast comparisons, it would equal or outperform the forecasting skill of both model 1 and model 2 (as guaranteed by the theorem of @Gneiting2007), while also matching the economic utility of model 1 (as guaranteed by the theorem of @Reed1979).
In practice, we never have access to the generating model, so it is reasonable to expect model selection to determine the better approximation.
As we see here, the better approximation for forecasting future states does not in fact lead to better outcomes.

@@

In practice, managers rarely rely only on forecasting skill to assess models, nor determine policies directly from forecasts alone.
In both examples presented above, the forecast trap is most likely in circumstances where the collection of candidate models is insufficiently broad.
Management practice tends to emphasize approaches which _broaden_ rather than _narrow down_ this candidate set.
This reflects the view that "the primary values of ecosystem models are as heuristic tools for communication and for developing scenarios to express uncertainties and test policies" rather than as a source of reliable forecasts [@Schindler2015].
The practices include:

@@

Second, A manager may also escape the forecast trap by exploring actions which are never recommended from any of the available forecasts.
For example, a manager looking at the low stocks and poor harvests achieved under model 2, could decide experimentally to reduce fishing quota. 
This will allow the population to enter a range of state-space where discrepancies between model 2 and observations are more obvious.
@Runge2016 describes how such "double-loop learning" can be used to identify when the entire model set is inaccurate, a problem which is not solved by "single-loop" adaptive-management in our examples above.
@Schindler2015 also underscores the value of flexible policies; rather than "managing solely within the range of past variation; active probing is usually needed," and contrasts this to a typical interpretation of the 'precautionary principle' often cited as a reason to avoid exploratory actions.
However, just because active exploration can escape the forecast trap does mean it is always a good idea.

Third, managers may emphasize policy robustness over forecast skill [@Schindler2015]. 
In many formal treatments, this is not qualitatively different to the analysis considered here: 
a manager simply chooses a different utility function, such as minimizing 'regret' rather than maximizing expected value [@Polasky2011]. 
Such approaches are just as vulnerable to bad outcomes (as defined by their own utility functions) whenever models are selected only on the basis of forecast skill.
But in practice, robust design may emphasize acceptable performance across the widest possible array of scenarios (e.g. candidate models).
This acts more like a sensitivity analysis of utility with respect to underlying assumptions, rather than an optimization routine [e.g. @Punt2016; @Fischer2009].
Computationally, the former is much simpler, allowing researchers to evaluate the performance of a policy on more complex simulations for which calculating the optimal policy would be prohibitively difficult.

>>>>

- abbreviated running title: "The Forecast Trap"
- Authorship statement: _This is a single-author paper_
- data accessibility statement: _All simulation data generated for analyses here, along with code required for the analysis, is available in the Zenodo Data archive, https://doi.org/10.5281/zenodo.4660621_
- number of words in the abstract: 148
- number of words in main text: 4950
- number of cited references: 51
- number of tables & figures: 5 Figures

@@

In this paper, I demonstrate that the model that makes the better prediction (defined as a strictly proper score, @Gneiting2007) is not necessarily the model that makes the better policy (defined in terms of utility, e.g. expected net present value, @Clark1990).
I show that our best methods for learning about model structure or parameters by repeatedly comparing forecasts to observations can be counter-productive.
Put another way, the value of information (VOI, as measured by the expected utility given that information minus the utility without it; see @Howard1966; @Katz1987), can actually be negative.
When VOI is negative, the decision-maker may become trapped into accepting mediocre outcomes derived from a model that makes accurate forecasts, even when a less accurate model that would generate better outcomes is available.
This trap is invisible to the manager unless sufficient alternative models outside the original set are introduced.
I will present two examples of this "forecast trap" and examine how it arises as a result of non-uniqueness of models [@Oreskes1994; @Schindler2015] with respect to either of these objectives.

@@

I shall focus on examples involving fisheries management to illustrate principles shared in many ecological systems.
Fisheries are a significant economic and conservation concern worldwide and their management remains an important debate [e.g. @Worm2006; @Worm2009; @Costello2016].
Moreover, fisheries management has been a proving grounds for theoretical and practical decision-making issues [e.g. @Clark1973; @Reed1979; @Walters1981; @Ludwig1982] arising in a wide range of other contexts, including
invasive species [@Boettiger2021], infectious diseases [@Li2017; Shea2014], fire management [@Richards1999] conservation planning and prioritization [@Chades2008; @Wilson2006] climate policy [@Nitzbon2017] and much else [@Ludwig1993; @Lande1994; @Polasky2011].

@@

To better understand how a model can produce a more accurate forecast and yet still lead to a worse decision, it may be helpful to start with a concrete example in which a manager faces a trade-off between cormorant conservation and fish harvest.
Fig 1 shows both the forecasts and realized management outcomes of two alternative three-species models, "A" and "B" (see Appendix A) in predicting the population dynamics of striped bass (an economically important fishery) and double-crested cormorants (a target species for conservation) which both feed on a population of river herring (whose abundance we assume is not measured). 
Model A accurately forecasts the abundance of both bass and cormorants well into the future, but the optimal management strategy derived from its forecasts leads to steadily declining species abundances and overall disappointing net utility.
Model B produces substantially less accurate forecasts, but nevertheless achieves better outcomes.
The net utility under model B is in fact nearly identical to the maximum utility attainable given the true model, while management under model A achieves only 38% of that utility.

@@

Faced with a collection of models, a manager can seek either to identify the best model to use, or to consider an integrative assessment which uses the whole ensemble of models to represent the manager's uncertainty about the underlying process.
I will consider both approaches in turn. 
Figure 2 compares forecasts generated by two of the candidate models to observations drawn from simulations of the underlying process.
As before, these are true forecasts: the model forecasts are generated first, they have not been fit to these observations.
In the un-fished scenario (top panels), both models try to predicting the same un-fished equilibrium dynamics.
In the second scenario (lower panel), the manager uses the optimal SDP policy derived from each forecast to determine the TAC for the following year, and compares the observed stock size to that which the model predicted given that fishing quota.
In both cases, model 2 provides far more accurate forecasts, as seen in the error bars and confirmed by the distribution of proper scores [Fig 2C-D; @Gneiting2007].
Model and simulation details are provided in Appendix B.

@@

Despite the clearly superior predictive accuracy of model 2 in both scenarios, the outcomes from management under model 2 are substantially worse.
We can assess such outcomes in less abstract terms than forecasting skill, such as economic value them manager sought to optimize (in dollars) or the ecological value (unharvested biomass) [Fig 3].

@@

A manager operating under model 2 would have little indication that the model was flawed: both future stock sizes and expected harvest yields consistently match model predictions.
This manager would be stuck in the forecast trap, incorrectly mistaking a degraded ecological state and reduced economic outcomes as the best that can be achieved because future observations continue to validate this model.
Had we been able to include Model 3 in our forecast comparisons, it would equal or outperform the forecasting skill of both model 1 and model 2 (as guaranteed by the theorem of @Gneiting2007), while also matching the economic utility of model 1 (as guaranteed by the theorem of @Reed1979).
In practice, we never have access to the generating model, so it is reasonable to expect model selection to determine the better approximation.
As we see here, the better approximation for forecasting future states does not in fact lead to better outcomes.

@@

In practice, managers rarely rely only on forecasting skill to assess models, nor determine policies directly from forecasts alone.
In both examples presented above, the forecast trap is most likely in circumstances where the collection of candidate models is insufficiently broad.
Management practice tends to emphasize approaches which _broaden_ rather than _narrow down_ this candidate set.
This reflects the view that "the primary values of ecosystem models are as heuristic tools for communication and for developing scenarios to express uncertainties and test policies" rather than as a source of reliable forecasts [@Schindler2015].
Such practices include:

@@

Second, a manager may also escape the forecast trap by exploring actions which are never recommended from any of the available forecasts.
For example, a manager looking at the low stocks and poor harvests achieved under model 2, could decide experimentally to reduce fishing quota. 
This will allow the population to enter a range of state-space where discrepancies between model 2 and observations are more obvious.
@Runge2016 describes how such "double-loop learning" can be used to identify when the entire model set is inaccurate, a problem which is not solved by "single-loop" adaptive-management in our examples above.
@Schindler2015 also underscores the value of flexible policies; rather than "managing solely within the range of past variation; active probing is usually needed," and contrasts this to a typical interpretation of the 'precautionary principle' often cited as a reason to avoid exploratory actions.
However, just because active exploration can escape the forecast trap does not mean it is always a good idea.

Third, managers may emphasize policy robustness over forecast skill [@Schindler2015]. 
In many formal treatments, this is not qualitatively different to the analysis considered here: 
a manager simply chooses a different utility function, such as minimizing 'regret' rather than maximizing expected value [@Polasky2011]. 
Such approaches are just as vulnerable to bad outcomes (as defined by their own utility functions) whenever models are selected only on the basis of forecast skill.
Alternative approaches may not seek any such optimization, emphasize the viability [@Aubin1991] of possible policy under constraints.
In practice, robust design may emphasize acceptable performance across the widest possible array of scenarios (e.g. candidate models).
This acts more like a sensitivity analysis of utility with respect to underlying assumptions, rather than an optimization routine [e.g. @Punt2016; @Fischer2009].
Computationally, the former is much simpler, allowing researchers to evaluate the performance of a policy on more complex simulations for which calculating the optimal policy would be prohibitively difficult.
